How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-02-25 doc/paper.tex
updating graphs. everything runs headless. table of model desc
seth comments on manuscript filtered through. supp info. fewer figures. regen figures.

\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{microtype, parskip, graphicx}
\usepackage[comma,numbers,sort&compress]{natbib}
\usepackage{lineno}
\usepackage{longtable}
\usepackage{docmute}
\usepackage{caption, subcaption, multirow, morefloats, rotating}
\usepackage{wrapfig}
\usepackage{hyperref}

@@

@@

\begin{abstract}
  One of the great promises of paleobiology is that by studying the past we can better predict the future.  This promise is particularly pertinent given as risk assessments for some modern species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.  Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)? (2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?  
  To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). We analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature. %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future. 
  Our results show that our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction. We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance. 
  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. The greatest source of variance in survival probability is the timing of observation. Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
  We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.  These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

@@

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval (Fig. \ref{fig:temp_curve}). Temperature was then transformed and standardized the in the same manner as geographic range (above). Difference in temperature between between observations was measured from the transformed and standardized temperature values and not standardized seperately.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cramer_temp}
  \caption{Comparison of initial temperature estimates from \citet{Cramer2011} (goldenrod) versus the binned values used in this analysis (blue). The initial values are for every 0.1 My while our bins are defined for every 1 My.}
  \label{fig:temp_curve}
\end{figure}

@@

@@

In total, our model can be expressed as: 
\begin{equation}
  \begin{aligned}
    t_{i} &\sim \text{Bernoulli}(\Theta) \\
    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{l[i], p[i]}) \\
    B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
    \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
    % double check this\dots is A MVN dist?
    A_{l, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
    \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
    \mu_{d} &\sim 
      \begin{cases}
        N(-2, 5) & \text{if } d = \text{intercept} \\
        N(-1, 1) & \text{if } d = \text{geo. range} \\
        N(0, 1) & \text{else } \\
      \end{cases} \\
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
    \tau_{B} &\sim C^{+}(5) \\
    \tau_{\alpha} &\sim C^{+}(5) \\
    \tau_{A} &\sim C^{+}(5) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
    \Omega_{A} &\sim LKJ(1) \\
  \end{aligned}
  \label{eq:priors}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.

We considered four variants of the model described above: 1) historical covariates with time-varying intercepts and slopes, 2) no historical coviarates with time-varying intercepts and slopes, 3) historical covariates without time-varying intercepts and slopes, and 4) no historical covariates and no time-varying intercepts and slopes (except age effect). The second and fourth models modify the number of columns in \(X\) by removing two of the covariates. The third and fourth model simplify the first model to just varying intercepts without varying slopes.

\subsection{Model Parameter Estimation}

We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language \citep{StanManual}. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.

%An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
%\begin{verbatim}
%form <- event ~ range + range_diff + temp + temp_lag + 
%                (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
%                (1 | age/phylum), 
%stan_glmer(formula = form,
%           data = data, 
%           family = 'binomial',
%           prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
%           prior_intercept = normal(0, 10, autoscale = FALSE), 
%           prior_aux = cauchy(0, 3, autoscale = FALSE), 
%           chains = 4,
%           thin = 4,
%           adapt_delta = 0.999999)
%\end{verbatim}
%
%Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} \citep{brms2017,brms2018} as:
%\begin{verbatim}
%priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
%            set_prior('normal(0, 1)', class = 'b'),
%            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
%            set_prior('cauchy(0, 3)', class = 'sd'),
%            set_prior('lkj(1)', class = 'cor'))
%form <- bf(event ~ range + range_diff + temp + temp_lag +
%           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
%           (1 | age/phylum))
%brmfit <- brm(formula = form,
%              data = data, 
%              family = bernoulli(), 
%              prior = priors,
%              chains = 4, 
%              thin = 4,
%              control = list(adapt_delta = 0.999999)
%\end{verbatim}

@@

The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to. Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig. \ref{fig:auc_hist}). The parameter rich ``past and vary'' model has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich ``past and vary'' model is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance. It is then hard to conclude that there is one ``best'' model which we can rely upon.

% ROC model comparison
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist}
  \caption{Posterior predictive AUC estimates for each of the four models being compared. These estimates are calculated from each of the models posterior predictive distribution compared to the empirical values. Models with a higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1.}
  \label{fig:auc_hist}
\end{figure}

When the posterior predictive distributions of the in-sample AUC estimates are presented over time, the similarity in adequacy between the models becomes more apparent (Fig. \ref{fig:auc_ts}). There are few major or obvious differences in model adequacy between the four models.
% ROC model comparison time series
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_ts}
  \caption{Comparison between the posterior predictive AUC estimates for each of the time intervals for each of the four models. These estimates are reflections of each model's fit to the various time intervals. The red line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
  \label{fig:auc_ts}
\end{figure}

When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig. \ref{fig:auc_taxon}). While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig. \ref{fig:auc_hist}). For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria. 
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon}
  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups present in this analysis. The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
  \label{fig:auc_taxon}
\end{figure}

>>>>

\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{microtype, parskip, graphicx}
\usepackage[comma,numbers,sort&compress]{natbib}
\usepackage[margin=1in]{geometry}
\usepackage{lineno}
\usepackage{longtable}
\usepackage{caption, subcaption, multirow, morefloats, rotating}
\usepackage{wrapfig}
\usepackage{hyperref}
\usepackage{threeparttable}

@@

\newcommand{\beginsupplement}{
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S\arabic{equation}}
}

@@

\begin{abstract}
  One of the great promises of paleobiology is that by studying the past we can better predict the future. This promise is particularly pertinent given as risk assessments for some modern species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales. Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)? (2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?  
  To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). 
  We analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature. 
  %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future. 
  Our results show that our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. 
  We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance. 
  Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
  However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction. 
  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. 
  %This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. 
  %The greatest source of variance in survival probability is the timing of observation. 
  %Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
  %We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. 
  %The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. 
  These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.  
  These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
  The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

@@

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval. Temperature was then transformed and standardized the in the same manner as geographic range (above). Difference in temperature between between observations was measured from the transformed and standardized temperature values and not standardized seperately.

%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cramer_temp}
%  \caption{Comparison of initial temperature estimates from \citet{Cramer2011} (goldenrod) versus the binned values used in this analysis (blue). The initial values are for every 0.1 My while our bins are defined for every 1 My.}
%  \label{fig:temp_curve}
%\end{figure}

@@

We developed a discrete-time survival model to analyze our data in order to answer our question of how wee can we predict extinction risk at million year time scales. We considered four model variations: covariate effects are constant over time and none of our historical covariates are included (Model C), covariate effects are allowed to vary over time but we include none of our historical covariates (Model V), covariate effects are constant over time and our historical covariates are included (Model CP), covariate effects are allowed to vary over time and we include our historical covariates (Model VP). See Table \ref{tab:model_def} for further explanation of how these models differ from each other. For a complete description of the statistical model used in this analysis, please see Section \ref{sec:model_desc}. For a description of how our models were implemented, please see Section \ref{sec:model_est}.

\begin{table}[ht]
  \caption{Models and their definitions}
  \begin{threeparttable}
    {
      \def\arraystretch{1.5}
      \begin{tabular}{ l p{3cm} l l }
        Code & Description & Covariates & R Formula Syntax\tnote{a}\phantom{\textsuperscript{a}} \\
        \hline
        C & Constant effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event\tnote{b}\phantom{\textsuperscript{b}} $\sim$ range\tnote{c}\phantom{\textsuperscript{c}} + temp\tnote{d}\phantom{\textsuperscript{d}} +\\(1 $|$ age\tnote{e}\phantom{\textsuperscript{e}}/phylum\tnote{f}\phantom{\textsuperscript{f}})} \\
        V & Varying effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ phylum) + (1 $|$ age/phylum)} \\ 
        CP & Constant effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff\tnote{g}\phantom{\textsuperscript{g}} +\\temp + temp\_lag\tnote{h}\phantom{\textsuperscript{h}} +\\(1 $|$ age/phylum)} \\
        VP & Varying effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff +\\temp + temp\_lag +\\(1 + range + range\_diff +\\temp + temp\_lag $|$ phylum) +\\(1 $|$ age/phylum)} \\
        \hline
      \end{tabular}
    }
    \begin{tablenotes}
    \item[a] See Equation \ref{eq:model} for full statistical model definition.
    \item[b] Species observation where 1 if time of last observation, otherwise 0.
    \item[c] Species geographic range in log km\(^2\). Mean centered, scaled to sd = 1.
    \item[d] Global temperature in degrees C. Mean centered, scaled to sd = 1.
    \item[e] Species are at observation in millions of years.
    \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcaeous nannoplankton).
    \item[g] Change in geographic range since last observation.
    \item[h] Temperature at previous observation.
    \end{tablenotes}
  \end{threeparttable}
  \label{tab:model_def}
\end{table}

% ROC model comparison
\begin{figure}[ht]
  \begin{subfigure}[ht]{0.45\textwidth}
    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
    \caption{In-sample}
    \label{fig:auc_hist}
  \end{subfigure}
  \begin{subfigure}[ht]{0.45\textwidth}
    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_full}
    \caption{Out-of-sample}
    \label{fig:fold_auc}
  \end{subfigure}
  \caption{In-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) AUC estimates for each of our four models. These estimates are calculated from the models posterior predictive distribution or from predictions made to new data, respectively. Models with a higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:auc_compare}
\end{figure}

The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to. Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig. \ref{fig:auc_hist}). The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77 which is interpreted as ``fine but not good'' performance. It is then hard to conclude that there is one ``best'' model which we can rely upon.

%When the posterior predictive distributions of the in-sample AUC estimates are presented over time, the similarity in adequacy between the models becomes more apparent (Fig. \ref{fig:auc_ts}). There are few major or obvious differences in model adequacy between the four models.
%% ROC model comparison time series
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_ts_full}
%  \caption{Comparison between the posterior predictive AUC estimates for each of the time intervals for each of the four models. These estimates are reflections of each model's fit to the various time intervals. The red line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
%  \label{fig:auc_ts}
%\end{figure}

%When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig. \ref{fig:auc_taxon}). While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig. \ref{fig:auc_hist}). For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria. 
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_full}
%  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups present in this analysis. The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
%  \label{fig:auc_taxon}
%\end{figure}

For many taxon/model combinations there are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 -- AUC value of 0.5 indicates that the model's predictions are no better than random (Fig. \ref{fig:auc_taxon_time}). However, this pattern is absent for the posterior predictive distribution of Foraminfera and Radiolaria for the VP model. Additionally, these periods of low model performance are rarer for the posterior predictive distribution of the VP model for calcareous nannoplankton and Dinoflagellates when compared to the other three models.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged over time and by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:auc_taxon_time}
\end{figure}

Comparison between the posterior predictive distributions of expected out-of-sample AUC (Fig. \ref{fig:fold_auc}) reveals a similar range in plausible values for all models as the in-sample AUC posterior predictive distributions (Fig. \ref{fig:auc_hist}). Interestingly, the differences between the posterior predictive distributions for the models have decreased. For example, model VP not clearly better than either models V or CP (Fig. \ref{fig:fold_auc}), which were shown earlier to be obviously worse-performing models based on in-sample performance (Fig. \ref{fig:auc_hist}). These differences means that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC. However, the shapes of the posterior distributions means interepting from the median values is incorrect -- the models are effectively indistinguishable in their expected out-of-sample AUC values.

Additionally, the quality of expected out-of-sample performance is not great, with average out-of-sample AUC for each of our models estimated to be between 0.7 and 0.8 which is far from perfect. This result means that we would expect to correctly rank two species in order of most to least likely to go extinct 70-80\% of the time. However, this expected out-of-sample performance is approximately the same as the in-sample performance results (Fig. \ref{fig:auc_hist}), indicating that our models would yield consistent results when generalized to future extinctions.

%When the posterior predictive distribution of expected out-of-sample AUC is presented as a time series, the similarity between the models is even more apparent (Fig. \ref{fig:fold_auc_time}). While the width of the credible intervals at various time points varies between the models, the overall picture of expected out-of-sample AUC is almost identical when you compare the models -- periods of relatively better or worse performance map identically between the time series.
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time_full}
%  \caption{Comparison of out-of-sample AUC values calculated for each of the My intervals for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds. This presentation decomposes each of the 12-million year folds (Fig. \ref{fig:fold_auc}) into the predictions made for each of the million-year intervals. The red line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.}
%  \label{fig:fold_auc_time}
%\end{figure}
%We can also compare expected out-of-sample AUC by taxonomic group for each of the models (Fig. \ref{fig:fold_auc_taxon}). These comparisons reveal a lot about the differences in predictive potential of the taxonomic groups. For example, the posterior predictive distributions of out-of-sample AUC for Foraminifera from all four models are approximately identical. In contrast, expected out-of-sample AUC for Radiolaria exhibits the same or similar pattern in relative model performance to the pooled comparisons (Fig. \ref{fig:fold_auc}). Additionally, we can state that our out-of-sample predictions for calcareous nannoplankton and dinoflagellates are not necessarily as precise as our estimates for Foraminifera or even Radiolaria. These results indicate that out-of-sample predictions may be easier for some taxonomic groups than others (e.g. Foraminifera versus Dinoflagellates.
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_full}
%  \caption{Comparison of out-of-sample AUC values aggregated by taxonomic group for each of the four models. Depicted for each taxon-model combination is an aggregate density of all posterior predictive estimates for each of four folds -- cross-validation estimates are commonly multi-modal as each fold presents its own challenges for prediction. Beneath these densities is marked the median estimate along with 50\% and 80\% credible intervals.}
%  \label{fig:fold_auc_taxon}
%\end{figure}
%

We can also compare the posterior predictive distribution of expected out-of-sample AUC over time and taxonomic group for each of the four models (Fig. \ref{fig:fold_auc_taxon_time}). For each taxonomic group, the time-series of posterior predictive values for each model are broadly congruent.

In the analysis of the posterior predictive distributions of the in-sample AUC values for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior predictive distribution of out-of-sample AUC values -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random. In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with model V.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
  \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:fold_auc_taxon_time}
\end{figure}

One of the most striking results of this analysis is that the in-sample and out-of-sample comparisons between our models demonstrate that while models where the historical coarviates are predictors of species extinction have better in-sample performance (Fig. \ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them generalize to out-of-sample data with a similar degree of success (Fig. \ref{fig:fold_auc}, \ref{fig:fold_auc_taxon_time}).

Other noteworthy aspects of our in-sample results are that AUC estimates for our models models differ significantly between models and that all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible (Fig. \ref{fig:fold_auc}). This means that even though model VP has a \textit{statistically} greater AUC than the other three models, this result is not practically or \textit{scientifically} significant.

\section{Supplementary Material}
\beginsupplement

\subsection{Model Specifications} \label{sec:model_desc}

@@

In total, our model can be expressed as: 
\begin{equation}
  \begin{aligned}
    t_{i} &\sim \text{Bernoulli}(\Theta) \\
    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{l[i], p[i]}) \\
    B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
    \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
    % double check this\dots is A MVN dist?
    A_{l, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
    \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
    \mu_{d} &\sim 
    \begin{cases}
      N(-2, 5) & \text{if } d = \text{intercept} \\
      N(-1, 1) & \text{if } d = \text{geo. range} \\
      N(0, 1) & \text{else } \\
    \end{cases} \\
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
    \tau_{B} &\sim C^{+}(1) \\
    \tau_{\alpha} &\sim C^{+}(1) \\
    \tau_{A} &\sim C^{+}(1) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
    \Omega_{A} &\sim LKJ(1) \\
  \end{aligned}
  \label{eq:model}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.

\subsection{Model Parameter Estimation} \label{sec:model_est}

We implemented our model (Eq. \ref{eq:model} using the \begin{texttt}rstanarm\end{texttt} package for the R programming language \citep{StanManual}. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.

An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
\begin{verbatim}
form <- event ~ range + range_diff + temp + temp_lag + 
                (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                (1 | age/phylum), 
stan_glmer(formula = form,
           data = data, 
           family = 'binomial',
           prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
           prior_intercept = normal(-2, 5, autoscale = FALSE), 
           prior_aux = cauchy(0, 1, autoscale = FALSE), 
           chains = 4,
           thin = 4,
           adapt_delta = 0.999999)
\end{verbatim}

Similarly, our full model can also be implemented using the \begin{texttt}brms\end{texttt} Stan interface \citep{brms2017,brms2018} as:
\begin{verbatim}
priors <- c(set_prior('normal(-2, 5)', class= 'Intercept'),
            set_prior('normal(0, 1)', class = 'b'),
            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
            set_prior('cauchy(0, 1)', class = 'sd'),
            set_prior('lkj(1)', class = 'cor'))
form <- bf(event ~ range + range_diff + temp + temp_lag +
           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
           (1 | age/phylum))
brmfit <- brm(formula = form,
              data = data, 
              family = bernoulli(), 
              prior = priors,
              chains = 4, 
              thin = 4,
              control = list(adapt_delta = 0.999999)
\end{verbatim}

@@
